---
title: Using the `ResidualMatrix` class
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com 
date: "Revised: 30 August 2020"
output:
  BiocStyle::html_document:
    toc: false
package: ResidualMatrix
vignette: >
  %\VignetteIndexEntry{Using the ResidualMatrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
set.seed(100)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Overview

A common step in genomics involves computing residuals to regress out uninteresting factors of variation.
However, doing so naively would discard aspects of the underlying matrix representation.
The most obvious example is the loss of sparsity when a dense matrix of residuals is computed,
increasing memory usage and compute time in downstream applications.
This motivates the use of the `ResidualMatrix` class, which provides an alternative to explicit calculation of the residuals.

# Using the `ResidualMatrix`

The constructor takes a matrix of input values and a design matrix, 
where residuals are conceptually computed by fitting the linear model to the columns of the input matrix.
However, the actual calculation of the residuals is delayed until they are explictly required.

```{r}
design <- model.matrix(~gl(5, 10000))

# Making up a large-ish sparse matrix.
library(Matrix)
y0 <- rsparsematrix(nrow(design), 30000, 0.01)

library(ResidualMatrix)
resids <- ResidualMatrix(y0, design)
resids
```

It is simple to obtain the residuals for, say, a single column.
We could also use the `r Biocpkg("DelayedArray")` block processing machinery to do this for chunks of columns at a time,
allowing downstream code to compute on the residuals within memory limits.

```{r}
hist(resids[,1])
```

In fact, matrix multiplication steps involving a `ResidualMatrix` do not need to compute the residuals at all.
This means that `ResidualMatrix` objects can be efficiently used in approximate PCA algorithms based on multiplication - 
see the `r Biocpkg("BiocSingular")` package for some examples.
The only requirement is that the original matrix has a reasonably efficient matrix multiplication operator.

```{r}
library(rsvd)
set.seed(100)
system.time(pc.out <- rsvd(resids, 10)) 
str(pc.out)
```

Similarly, the row and column sums/means can be computed efficiently, 
based on the matrix multiplication machinery and the original matrix's row and column sum functions.

```{r}
hist(rowSums(resids))
```

Other operations will cause the `ResidualMatrix` to collapse into `DelayedMatrix` for further processing.

# Retaining certain factors

We can also specify that we only want to regress out some factors in our `design`.
For example, let's say we have a dataset with an interesting two-group structure and an uninteresting continuous covariate `BAD`:

```{r}
design2 <- model.matrix(~gl(2, 10000))
design2 <- cbind(design2, BAD=runif(nrow(design2)))
colnames(design2)
```

We can instruct `ResidualMatrix()` to retain the interesting structure (first two coefficients)
while regressing out the uninteresting covariate in the third coefficient:

```{r}
# Making up another large-ish sparse matrix.
y0 <- rsparsematrix(nrow(design2), 30000, 0.01)

resid2 <- ResidualMatrix(y0, design2, keep=1:2)
resid2
```

In this sense, the `ResidualMatrix` is effectively a delayed version of `removeBatchEffect()`,
the old workhorse function from `r Biocpkg("limma")`.

# Session information

```{r}
sessionInfo()
```
